Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-12-09 paper/manuscript.tex

To derive an RBFE formulation from double-decoupling, we write expressions for the binding constants of $A$ and $B$ using Eq.~(\ref{eq:ddm-ratios-1}) and taking their ratio. While doing so, the standard state, pure solvent, and gas-phase terms cancel, and we obtain
\begin{equation}\label{eq:double-transformation-Kb-rbfe}
\frac{K_b(B)}{K_b(A)} = \frac{Z_{N_v,RB}}{Z_{N_v,RA}} \frac{Z_{N_v,A}}{Z_{N_v,B}}
\end{equation}
where the first ratio of partition functions involves receptor-ligand complexes and the second only the ligands in solution. Eq.~(\ref{eq:double-transformation-Kb-rbfe}), suggests that the relative binding free energy, $\Delta \Delta G_b$, of $B$ relative to $A$ can be expressed as the difference of the free energies of ``mutating'' ligand $A$ into ligand $B$ in the presence and absence of the receptor (Figure \ref{fig:fig_binding_thermodynamic_cycle}):
\begin{equation}\label{eq:double-transformation-dg-rbfe}
\Delta \Delta G_b = -k_B T \ln \frac{Z_{N_v,RB}}{Z_{N_v,RA}} - (-k_B T \ln \frac{Z_{N_v,B}}{Z_{N_v,A}}) = \Delta G_{\rm bound} - \Delta G_{\rm unbound}
\end{equation}
In practice, each ratio of partition functions must first be turned into an ensemble average to evaluate it. Similar to the preceding case, this step requires manipulating the integrals so that they are carried over the same set of degrees of freedom. As discussed in Section \ref{sec:relative-fe-protocol}, this step involves using alchemical topologies capable of describing both ligands. For example, among other features, alchemical topologies include atoms common to both ligands and dummy atoms that exist in one but not in the other ligand (Figure \ref{fig:fig_topology}). In one limit, called dual coordinate dual topology in Section \ref{sec:relative-fe-protocol}, all atoms of both ligands are present, and the partition functions in Eq.~(\ref{eq:double-transformation-dg-rbfe}) differ in which ligand is coupled to the system. A key aspect of constructing a suitable alchemical topology is the requirement that the additional degrees of freedom do not affect the RBFE estimate.

@@

The RBFE alchemical transfer formulation is very similar to the ABFE formulation. The simulation system includes both ligands, one in solution and the other in the binding site, and the coordinate transformation involves displacing them in opposite directions so that they switch places. Formally,~\cite{azimi2022relative} the method derives from Eq.~(\ref{eq:DG0-binding-constant-def}) for each binding constant
\begin{equation}
\frac{K_b(B)}{K_b(A)} = \frac{z_{RB(v)} z_{A(v)}}{z_{B(v)} z_{RA(v)}}
\end{equation}
and noting that the numerator represents a system with $B$ bound and $A$ unbound, and in the denominator, the roles are switched. Next, the solvent PMF is used to indicate the solvent degrees of freedom explicitly, and the indicator functions $I^\ast(\zeta_A)$ and $I^\ast(\zeta_b)$ are introduced to keep the unbound ligands in the solvent, similarly to Eq.~(\ref{eq:atm-partfunc-ratio}). The result is
\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-1}
  \frac{K_b(B)}{K_b(A)} = \frac{
    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I^\ast(\zeta_A) I(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
  }{
    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
  }
\end{equation}
where $x_A$ and $x_B$ are the internal coordinates of the ligands and $\zeta_A$, $\zeta_B$ are their external coordinates relative to the receptor.  Next we apply the change of variables $\zeta_B \rightarrow \zeta_B + \vec{h}$ and $\zeta_A \rightarrow \zeta_A - \vec{h}$ to the partition function in the numerator to obtain
\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-2}
  \frac{K_b(B)}{K_b(A)} = \frac{
    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A + \vec{h}, x_B, \zeta_B - \vec{h}, \vec{r}_v)}
  }{
    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
  }
\end{equation}
which can be written as an ensemble average
\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-average}
\frac{K_b(B)}{K_b(A)} = \langle e^{-\beta \Delta U}\rangle_{RA+B}
\end{equation}
where
\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-pert}
\Delta U = U(x_R, x_A, \zeta_A + \vec{h}, x_B, \zeta_B - \vec{h}, \vec{r}_v) - U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)
\end{equation}
is the perturbation energy for displacing the two ligands in opposite directions, and the ensemble average $\langle \ldots \rangle_{RA+B}$ is carried out with $A$ bound and $B$ in the solvent. Except for the coordinate displacement, Eq.~(\ref{eq:atm-partfunc-ratio-rbfe-average}) is implemented similarly to the alchemical transfer ABFE protocol, including the use of the symmetric alchemical intermediate as in Eq.~(\ref{eq:atm-average-abfe-intermediate}).

>>>>

To derive an RBFE formulation from double-decoupling, we write expressions for the binding constants of $A$ and $B$ using Eq.~(\ref{eq:ddm-ratios-1}) and taking their ratio. While doing so, the standard state, pure solvent, and gas-phase terms cancel, and we obtain
\begin{equation}\label{eq:double-transformation-Kb-rbfe}
\frac{K_b(B)}{K_b(A)} = \frac{Z_{N_v,RB}}{Z_{N_v,RA}} \frac{Z_{N_v,A}}{Z_{N_v,B}}
\end{equation}
where the first ratio of partition functions involves receptor-ligand complexes and the second only the ligands in solution. Eq.~(\ref{eq:double-transformation-Kb-rbfe}), suggests that the relative binding free energy, $\Delta \Delta G_b$, of $B$ relative to $A$ can be expressed as the difference of the free energies, $\Delta G_{\rm bound}$ and $\Delta G_{\rm unbound}$, of ``mutating'' ligand $A$ into ligand $B$ in the presence and absence of the receptor, respectively (Figure \ref{fig:fig_binding_thermodynamic_cycle}):~\cite{tembre1984ligandreceptor,gilson1997statisticalthermodynamic,mobley2012perspective,cournia2017relative,jiang2019computing,lee_jcheminfmodel_2020_v60_p5595}
\begin{equation}\label{eq:double-transformation-dg-rbfe}
\Delta \Delta G_b = -k_B T \ln \frac{Z_{N_v,RB}}{Z_{N_v,RA}} - (-k_B T \ln \frac{Z_{N_v,B}}{Z_{N_v,A}}) = \Delta G_{\rm bound} - \Delta G_{\rm unbound}
\end{equation}

In practice, each partition function ratio must first be converted into an ensemble average to be evaluated. Similar to the preceding case, this step requires manipulating the integrals so that they are carried over the same set of degrees of freedom. As discussed in Section \ref{sec:relative-fe-protocol}, this step involves using alchemical topologies that can describe both ligands. For example, among other features, alchemical topologies include atoms common to both ligands and dummy atoms that exist in one but not in the other ligand (Figure \ref{fig:fig_topology}). A key aspect of constructing a suitable alchemical topology is the requirement that the additional degrees of freedom do not affect the RBFE estimate.

In the dual coordinate dual topology method (discussed in Section \ref{sec:relative-fe-protocol}), which is helpful for pairs of compounds that do not share the same scaffold, all atoms of both ligands are present but intermolecular interactions between the two ligands are turned off, and the partition functions in Eq.~(\ref{eq:double-transformation-dg-rbfe}) differ in which ligand is coupled to the system. An example of this formulation is the Separated Topologies (SepTop) approach~\cite{rocklin2013separated,baumann2023broadening}, which decouples one ligand as the coupling of the second is turned on. Additionally, the SepTop approach can estimate the relative binding free energies of pairs of ligands that do not share the same binding pose by employing different binding site indicator functions in the definitions of the partition functions of the complexes, $Z_{N_v,RA}$ and $Z_{N_v,RB}$, in Eq.~\ref{eq:double-transformation-dg-rbfe}.

@@

The RBFE alchemical transfer formulation follows closely the ABFE formulation. Similar to the dual coordinate dual topology method, the simulation system includes both ligands, one in solution and the other in the binding site, and the coordinate transformation involves displacing them in opposite directions so that they switch places. Formally,~\cite{azimi2022relative} the method derives from Eq.~(\ref{eq:DG0-binding-constant-def}) for each binding constant
\begin{equation}
\frac{K_b(B)}{K_b(A)} = \frac{z_{RB(v)} z_{A(v)}}{z_{B(v)} z_{RA(v)}}
\end{equation}
and noting that the numerator represents a system with $B$ bound and $A$ unbound, and in the denominator, the roles are switched. Next, the solvent PMF is used to indicate the solvent degrees of freedom explicitly, and the indicator functions $I^\ast(\zeta_A)$ and $I^\ast(\zeta_b)$ are introduced to keep the unbound ligands in the solvent, similarly to Eq.~(\ref{eq:atm-partfunc-ratio}). The result is
\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-1}
  \frac{K_b(B)}{K_b(A)} = \frac{
    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I^\ast(\zeta_A) I(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
  }{
    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
  }
\end{equation}
where $x_A$ and $x_B$ are the internal coordinates of the ligands and $\zeta_A$, $\zeta_B$ are their external coordinates relative to the receptor.  Next we apply the change of variables $\zeta_B \rightarrow \zeta_B + \vec{h}$ and $\zeta_A \rightarrow \zeta_A - \vec{h}$ to the partition function in the numerator to obtain
\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-2}
  \frac{K_b(B)}{K_b(A)} = \frac{
    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A + \vec{h}, x_B, \zeta_B - \vec{h}, \vec{r}_v)}
  }{
    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
  }
\end{equation}
which can be written as an ensemble average
\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-average}
\frac{K_b(B)}{K_b(A)} = \langle e^{-\beta \Delta U}\rangle_{RA+B}
\end{equation}
where
\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-pert}
\Delta U = U(x_R, x_A, \zeta_A + \vec{h}, x_B, \zeta_B - \vec{h}, \vec{r}_v) - U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)
\end{equation}
is the perturbation energy for displacing the two ligands in opposite directions, and the ensemble average $\langle \ldots \rangle_{RA+B}$ is carried out with $A$ bound and $B$ in the solvent. Except for the coordinate displacement, Eq.~(\ref{eq:atm-partfunc-ratio-rbfe-average}) is implemented similarly to the alchemical transfer ABFE protocol, including the use of the symmetric alchemical intermediate as in Eq.~(\ref{eq:atm-average-abfe-intermediate}).
